---
title: "coursework_Question2_R"
output: pdf_document
---

Question2: Do older planes suffer more delays?

```{r}
library(knitr)
```

Set working directory
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/juhyu/Desktop/Programming ST2195')
```

```{r}
# checking working directory
getwd()
```

Read dataset
```{r}
df2003<-read.csv("2003.csv.bz2")
df2004<-read.csv("2004.csv.bz2")
df2005<-read.csv("2005.csv.bz2")
df2<-rbind(df2003,df2004,df2005)
```

Extract even rows & selecting rows with flights not cancelled
```{r}
row_odd<-seq_len(nrow(df2)) %% 2
df2<-df2[row_odd==0,]
df2<-subset(df2,df2$Cancelled==0)
```

Read necessary libraries
```{r}
library(ggplot2)
library(lessR)
library(dplyr)
library(ggpubr)
```

Set necessary columns and read plane data
```{r}
df2<-df2[,c('Year',"DepTime","CRSDepTime",'TailNum','ArrDelay','DepDelay')]
planes<-read.csv('plane-data.csv')
```

Creating departure hour column
```{r}
class(df2$DepTime)
df2$DepHour<-floor(df2$DepTime/100)
df2<-subset(df2,df2$DepHour<=24)
```

```{r}
#checking null values and drop
table(is.na(planes))
planes<-na.omit(planes)

#sorting values to be year<2005
unique(planes$year)
planes<-subset(planes,year<2005)
```

Joining planes to df2 
```{r}
#Creating planes' age column
class(planes$year)
planes$age<-2022-as.numeric(planes$year)
length(planes$age)

#joining df to planes
names(planes)
names(planes)[1]<-'TailNum'
summary(planes$year)

df2<-merge(df2,planes,by='TailNum')
head(df2,5)
names(df2)
```

Plot box plot of age range
```{r}
#plot box plot
boxplot(df2$age,ylab="age")
summary(df2$age) #outlier: 2022
df2<-subset(df2,age<2022)
```

Create age group column
```{r}
labs<-c(paste(seq(10,60,by=10),seq(20,70,by=10),sep='-'))
df2$agegroup<-cut(df2$age,breaks=c(seq(10,60,by=10),Inf),labels=labs,right=FALSE)
head(select(df2,age,agegroup))
```

Plot bar plots of distribution of delays 
```{r}
#plotting distribution of total mean delay over a day 
mean_depdelay<-list()
mean_arrdelay<-list()

for (x in sort(unique(df2$agegroup))) {
  print(x)
  m_depdelay<-colMeans(df2[df2$agegroup==x,]['DepDelay'],na.rm=TRUE)
  mean_depdelay<-append(mean_depdelay,m_depdelay)
  m_arrdelay<-colMeans(df2[df2$agegroup==x,]['ArrDelay'],na.rm=TRUE)
  mean_arrdelay<-append(mean_arrdelay,m_arrdelay)
}
mean_depdelay
mean_arrdelay

deparr_delay<-mapply('+',mean_depdelay,mean_arrdelay)
deparr_delay<-data.frame(t(data.frame(deparr_delay)))
colnames(deparr_delay)<-sort(unique(df2$agegroup))
deparr_delay
Agegroup<-sort(unique(df2$agegroup))
Agegroup
```

To investigate deep down, we will plot the pie chart to compare each of their contribution to delay per different age groups. 
```{r}
#Plot bar chart
deparr_delay
barplot(as.matrix(deparr_delay),main='Distribution of delay by different ages of planes',xlab='Age of Planes',ylab='Delay(min)',col='#4F94CD')

legend(x="topright",inset=0.07,cex=0.8,
       pch=16,
       col=c('#11759D', '#FFA500'),
       legend = 'total mean delay per age',
       border = "black") # Color of the border of the squares
```

```{r}
#creating depdelay_perage
#depdelay_perage: total delay for each agegroup 
depdelay_perage<-list()
arrdelay_perage<-list()

for (x in sort(unique(df2$agegroup))) {
  print(x)
  depdelay_pa<-sum(df2[df2$agegroup==x,]['DepDelay'],na.rm=TRUE)
  depdelay_perage<-append(depdelay_perage,depdelay_pa)
  arrdelay_pa<-sum(df2[df2$agegroup==x,]['ArrDelay'],na.rm=TRUE)
  arrdelay_perage<-append(arrdelay_perage,arrdelay_pa)
}
depdelay_perage
arrdelay_perage

#turning depdelay_perage and arrdelay_perage into dataframe
depdelay_perage<-data.frame(t(data.frame(depdelay_perage)))
colnames(depdelay_perage)<-'delay'
rownames(depdelay_perage)<-Agegroup

arrdelay_perage<-data.frame(t(data.frame(arrdelay_perage)))
colnames(arrdelay_perage)<-'delay'
rownames(arrdelay_perage)<-Agegroup
```

```{r}
#plotting pie chart for depdelay_perage
depdelay_perage2<-depdelay_perage %>%
  mutate(perc=round((delay/sum(depdelay_perage$delay)),2)) %>% 
  mutate(prop=delay/sum(depdelay_perage$delay)*100) %>%
  mutate(ypos=cumsum(prop)-0.5*prop)%>%
  mutate(labels = scales::percent(perc))

depdelay_perage2$group<-Agegroup
rownames(depdelay_perage2)<-c(1:6)
depdelay_perage2

arrdelay_perage2<-arrdelay_perage %>%
  mutate(perc=round((delay/sum(arrdelay_perage$delay)),2)) %>% 
  mutate(prop=delay/sum(arrdelay_perage$delay)*100) %>%
  mutate(ypos=cumsum(prop)-0.5*prop)%>%
  mutate(labels = scales::percent(perc))

arrdelay_perage2$group<-Agegroup
rownames(arrdelay_perage2)<-c(1:6)
arrdelay_perage2
```


```{r}
#departure delay

pie_dep<-ggplot(depdelay_perage2, aes(x = "", y =delay, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y", start = 0) +
  scale_fill_manual(values = c("Blue", "Red", "Green", "Orange","Pink","LightBlue")) +
  labs(x = "", y = "", title = "Departure Delay",
       fill = "AgeGroup") + 
  geom_text(aes(x =1.9,y= delay, label =labels), color="black",
            fontface = "bold",position=position_stack(vjust=0.5)) +
  theme(plot.title = element_text(hjust = 0.5), 
        legend.title = element_text(hjust = 0.5, size = 5)) 

pie_dep

#arrival delay
pie_arr<-ggplot(arrdelay_perage2, aes(x = "", y =delay, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y", start = 0) +
  scale_fill_manual(values = c("Blue", "Red", "Green", "Orange","Pink","LightBlue")) +
  labs(x = "", y = "", title = "Arrival Delay",
       fill = "AgeGroup") + 
  geom_text(aes(x =1.9,y= delay, label =labels), color="black",
            fontface = "bold",position=position_stack(vjust=0.5)) +
  theme(plot.title = element_text(hjust = 0.5), 
        legend.title = element_text(hjust = 0.5, size = 5)) 

pie_arr

#combining them
ggarrange(pie_dep,pie_arr+rremove("x.text"),
          ncol=2,nrow=1,
          hjust=-2)
```



