---
title: "coursework_Question3_R"
output: pdf_document
---
Question3:How does the number of people flying between different locations change over time?

```{r}
library(knitr)
```

Set working directory
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/juhyu/Desktop/Programming ST2195')
```

```{r}
# checking working directory
getwd()
```

Read dataset
```{r}
df2003<-read.csv("2003.csv.bz2")
df2004<-read.csv("2004.csv.bz2")
df2005<-read.csv("2005.csv.bz2")
df3<-rbind(df2003,df2004,df2005)
```

Extract even rows & selecting rows with flights not cancelled
```{r}
row_odd<-seq_len(nrow(df3)) %% 2
df3<-df3[row_odd==0,]
df3<-subset(df3,df3$Cancelled==0)
```

Read necessary libraries 
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

str(df3)
table(df3$Year)
attach(df3)
```

We will measure number of people flying by number of flights. 
Creating barplot on number of flights each year
```{r}
flights_peryear<-data.frame(table(df3$Year))
names(flights_peryear)[1]<-'Year'
years<-c(2003,2004,2005)
colors<-'#5CACEE'
barplot(as.matrix(flights_peryear[2])~years,xlab="Year",col=colors)
```

To investigate further, we will look into trend in number of flights in the three years,in the top five airports
```{r}
table(df3$Origin)
flights_perairport<-data.frame(table(df3$Origin))
flights_perairport<-flights_perairport[order(flights_perairport$Freq,decreasing=TRUE),]
flights_perairport<-head(flights_perairport,5)
top5<-c('ATL','ORD','DFW','LAX','IAH')
```

Top 5 airports are :'ATL','ORD','DFW','LAX','IAH'
```{r}
head(df3,5)

#set df3 to contain only the data of top5 countries
df3<-filter(df3,df3$Origin%in%top5) 
unique(df3$Origin)
flights_summary<-df3%>%
  group_by(Year,Origin)%>%
  summarise(length(Year))
flights_summary
```

Creating data frame of flight information for three airports in 2003,2004,2005
```{r}
unique(df3$Year)
flights2003<-subset(flights_summary,flights_summary$Year=='2003')
flights2003<-t(flights2003[,c(2,3)])
flights2003<-flights2003[2,]
flights2003<-data.frame(t(data.frame(flights2003)))
colnames(flights2003)<-c('ATL','DFW','IAH','LAX','ORD')
```

```{r}
flights2004<-subset(flights_summary,flights_summary$Year=='2004')
flights2004<-t(flights2004[,c(2,3)])
flights2004<-flights2004[2,]
flights2004<-data.frame(t(data.frame(flights2004)))
colnames(flights2004)<-c('ATL','DFW','IAH','LAX','ORD')
```

```{r}
flights2005<-subset(flights_summary,flights_summary$Year=='2005')
flights2005<-t(flights2005[,c(2,3)])
flights2005<-flights2005[2,]
flights2005<-data.frame(t(data.frame(flights2005)))
colnames(flights2005)<-c('ATL','DFW','IAH','LAX','ORD')
```

```{r}
#create flights_info data frame
flights_info<-rbind(flights2003,flights2004,flights2005)
flights_info<-t(flights_info)

rownames(flights_info)
flights_info
```

Plotting bar graph based on number of flights in the top five airports throughout 3 years (2003,2004,2005)
```{r}
colnames(flights_info)<-c(2003,2004,2005)
flights_info<-data.frame(t(flights_info))
flights_info<-select(flights_info,ATL,ORD,DFW,LAX,IAH)
flights_info<-mutate_all(flights_info, function(x) as.numeric(as.character(x)))
```


```{r}
barplot(as.matrix(flights_info),
        xlab = "Airports",
        ylab = "Number of flights",
        beside=TRUE,
        col=c('skyblue3','coral','olivedrab1')
)
legend("top",
       c("2003","2004","2005"),
       col=c('skyblue3','coral','olivedrab1'),
       pch=16,
       cex=0.6
)
```
