---
title: "Question4_R"
output: pdf_document
---
Question4: Can you detect cascading failures as delays in one airport create delays in others? 

Reading necessary libraries
```{r}
library(dplyr)
library(ggplot2)
library(reshape2)
library(corrr)
library(tidyverse)
library(ggcorrplot)
library(knitr)
library(extrafont)
```

Set working directory
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/juhyu/Desktop/Programming ST2195')
```

```{r}
# checking working directory
getwd()
```

Read dataset
```{r}
df2003<-read.csv("2003.csv.bz2")
df2004<-read.csv("2004.csv.bz2")
df2005<-read.csv("2005.csv.bz2")
df4<-rbind(df2003,df2004,df2005)
```

Extract even rows & selecting rows with flights not cancelled
```{r}
row_odd<-seq_len(nrow(df4)) %% 2
df4<-df4[row_odd==0,]
df4<-subset(df4,df4$Cancelled==0)
```


```{r}
mydata<-df4[,c(1,2,3,4,5,6,7,8,12,13,14,15,16,19,24,25,26,27,28,29)]

names(mydata) #check if columns are added successfully
```

Creating correlation matrix
```{r}
cormat<-round(x=cor(mydata,use='complete.obs',method='pearson'),digits=2)
melted_cormat<-melt(cormat)
ggplot(data=melted_cormat,aes(x=Var1,y=Var2,fill=value))+
  geom_tile()+
  labs(fill="Pearson's\nCorrelation",title='Correlations in airline data')+
  geom_text(label=melted_cormat$value,size=2)+
  theme_classic()+
  theme(axis.text.x=element_text(angle=45,vjust=1,hjust=1))+
  scale_x_discrete(expand=c(0,0))+
  scale_y_discrete(expand=c(0,0))+
  theme(text=element_text(family ="Roboto"))
```
Correlation matrix for 'ArrDelay' and 'DepDelay' is high(0.9). 
Hence, we can assume that there might be a cascading failure between delays. 

We will now look for evidence of delay. 

Creating necessary columns and change column order
```{r}
#Creating date column
df4<-within(df4, Date <- sprintf("%d-%02d-%d", Year, Month, DayofMonth))

#Creating a total delay column
df4$totaldelay<-df4$DepDelay+df4$ArrDelay

#change column order
col_order<- c('Date','TailNum','CRSDepTime','DepTime','CRSArrTime','ArrTime','DepDelay','ArrDelay','totaldelay','Origin','Dest','LateAircraftDelay')
df4<-df4[,col_order]
head(df4,10)
#order them by date
df4%>%arrange(Date)
```


We will investigate rows with flights arriving at ATL airport on 2003-10-10 with arrival delay and rows with flights from ATL with departure delay. 
If scheduled departure timing was affected due to arrival delay around the same timing, cascading failure in delay is detected. 
```{r}
#Delayed arrival at ATL
head(filter(df4,df4$Dest=='ATL'& df4$ArrDelay>0 & df4$LateAircraftDelay>0 &
         df4$Date=='2003-10-10')%>%arrange(DepTime),7)

#Delayed departure from ATL
head(filter(df4,df4$Origin=='ATL'& df4$DepDelay>0 & df4$LateAircraftDelay>0 &
         df4$Date=='2003-10-10')%>%arrange(DepTime),7)
```
Late arrival of 'N105DA' flight caused departure delay of 'N826AS' flight. Presence of 'LateAircraftDelay' also proves cascading failure in delay.

Hence, there are cascading failures in delays. 



